---
title: "strawberries"
author: "Elizabeth Rosen"
format: html
editor: visual
---

## USDA NASS

```{r}
#| label: load libraries
#| warning: false
#| message: false

library(knitr)  
library(kableExtra)
library(tidyverse)
library(stringr)
```

## Read the file

```{r}
#| label: read data - glimpse 


strawberry <- read_csv("strawb_mar6.csv", 
                       col_names = TRUE,
                       show_col_types = FALSE)

source("my_functions.R")

```

Examine the data. How is it organized?

```{r}

strawb <- strawberry |> drop_one_value_col()


```

```{r}
#| label: explore strawb data

# assume data is a tibble
# n_show is the number of rows to show



show_unique <- function(data, nrows=10 ){
  # make a tibble items to hold the data to show
  # browser()
  a <- nrows * dim(data)[2]  # number of cells in items
  items <- rep(" ", a) # items will coerce everything to char
  dim(items) <- c(nrows ,dim(data)[2]) # shape items
  items <- as_tibble(items)
  colnames(items) <- colnames(data)
  # browser()
  for(i in 1:dim(data)[2]){

    col_items <- unique(data[,i])
    # row_ex is the number of rows needed 
    # to make the column length conformable with items
    row_ex <- nrows - dim(col_items)[1] 
    if(row_ex >= 0){
      ex_rows <- tibble(rep(" ",row_ex))
      colnames(ex_rows) <- colnames(col_items)
      col_add <- rbind2(col_items, ex_rows)
      
    } else if(row_ex < 0){
      col_add <- col_items[1:10,]
      
    }

    items[,i] <- col_add

  }
  
  return(items)
}

## test <- show_unique(strawb, 10)


```

```{r}
#|label: split strawb into census and survey pieces

strw_census <- strawb |> filter(Program == "CENSUS")

strw_survey <- strawb |> filter(Program == "SURVEY")

nrow(strawb) == (nrow(strw_census) + nrow(strw_survey))

```

```{r}
#| label: examine Census and survey tibbles
#| warning: false


s_census <- strw_census |> drop_one_value_col(prt_val = TRUE)

s_survey <- strw_survey |> drop_one_value_col(prt_val = TRUE)


unique_sur <- s_survey |> show_unique(nrows = 10)

unique_cen <- s_census |> show_unique(nrows = 10)


strw_census <- s_census |> select(-`State ANSI`)

strw_survey <- s_survey |> select(-`State ANSI`, -`Week Ending`, -Period)

rm(s_census, s_survey, strawberry, strawb, items)


```

```{r}
#| label: work on Census columns

commod <- strw_census$Commodity |> unique()

#### split Data Item

strw_census <- strw_census |>
  separate_wider_delim(  cols = Commodity,
                         delim = ",",
                         names = c("INCOME", 
                                   "NET CASH FARM",
                                   "STRAW"
                                               ),
                         too_many = "error",
                         names_sep = " ",
                         too_few = "align_start"
  )



inc <- strw_census$Fruit |> unique()

strw_census <- strw_census |>
  separate_wider_delim(  cols = Fruit,
                         delim = ",",
                         names = c("INCOME", 
                                   "STRAWB"
                                               ),
                         too_many = "error",
                         too_few = "align_start"
  )

```

```{r}
#| label: Assignment shortner results

straw_cen_f <- strw_census |> filter(State == "FLORIDA")

straw_sur_f <- strw_survey |> filter(State == "FLORIDA")
straw_cen_c <- strw_census |> filter(State == "CALIFORNIA")
straw_sur_c <- strw_survey |> filter(State == "CALIFORNIA")

rm(strw_census, strw_survey, unique_cen, unique_sur)

```

### Chemical usage

```{r}
#| warning: false
#| message: false
chem_categories_c <- straw_sur_c |> separate_wider_delim(cols='Domain Category',
                                                            delim=': ',
                                                            names=c('domain','chemical name'),
                                                            too_few='align_start',
                                                            too_many = 'error') %>% select(-domain)
chem_categories_c <- chem_categories_c |> separate_wider_delim(cols='chemical name',
                                                               delim='=',
                                                               names = c('name', 'number'),
                                                               too_few='align_start',
                                                               too_many='error') |> separate_wider_delim(cols='name', delim='(', names=c('extra', 'name'), too_few='align_start', too_many='error') %>% select(-extra)
unique_values_c <- unique(chem_categories_c$`name`)

chem_categories_f <- straw_sur_f |> separate_wider_delim(cols='Domain Category',
                                                            delim=': ',
                                                            names=c('domain','chemical name'),
                                                            too_few='align_start',
                                                            too_many = 'error') %>% select(-domain)
chem_categories_f <- chem_categories_f |> separate_wider_delim(cols='chemical name',
                                                               delim='=',
                                                               names = c('name', 'number'),
                                                               too_few='align_start',
                                                               too_many='error') |> separate_wider_delim(cols='name', delim='(', names=c('extra', 'name'), too_few='align_start', too_many='merge') %>% select(-extra)
unique_values_f <- unique(chem_categories_f$`name`)

```

# Let's look at the chemicals used in California and in Florida.


```{r}
#| warning: false
#| message: false
chems_c <- chem_categories_c %>%
  filter(name != "TOTAL)") %>%
  group_by(name) %>%
  summarise(n = n()) 

chems_f <- chem_categories_f %>%
  filter(name != "TOTAL)") %>%
  group_by(name) %>%
  summarise(n = n()) 

kable(chems_c, 
      col.names = c("Chemical", "Count"), 
      caption = "Count of Chemicals Used in California", 
      align = "lcc") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = F, 
    position = "center") %>%
  row_spec(0, bold = T, color = "white", background = "#0073e6") %>%
  row_spec(1:nrow(chems_c), background = "#f2f2f2") %>%
  column_spec(1, bold = T) %>%
  add_header_above(c("Chemical Information" = 2))

kable(chems_f, 
      col.names = c("Chemical", "Count"), 
      caption = "Count of Chemicals Used in Florida", 
      align = "lcc") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = F, 
    position = "center") %>%
  row_spec(0, bold = T, color = "white", background = "#e60000") %>%
  row_spec(1:nrow(chems_f), background = "#f9f9f9") %>%
  column_spec(1, bold = T) %>%
  add_header_above(c("Chemical Information" = 2))
```

Which ones do they have in common? Let's focus on these.

```{r}
#| warning: false
#| message: false
common_chems <- intersect(chems_c$name, chems_f$name)

california_chems <- chems_c %>%
  filter(name %in% common_chems) %>%
  group_by(name) %>%
  summarise(california_count = sum(n), .groups = "drop")  # Summing actual counts

# Step 3: Sum the occurrences for each chemical in Florida (chems_f)
florida_chems <- chems_f %>%
  filter(name %in% common_chems) %>%
  group_by(name) %>%
  summarise(florida_count = sum(n), .groups = "drop")  # Summing actual counts

# Step 4: Merge the counts from California and Florida on the chemical name
combined_chems <- full_join(california_chems, florida_chems, by = "name") %>%
  arrange(desc(california_count + florida_count))  # Sort by total count


kable(combined_chems, 
      col.names = c("Chemical", "California Count", "Florida Count"), 
      caption = "Counts of Common Chemicals in California and Florida", 
      align = "lcc") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = F, 
    position = "center") %>%
  row_spec(0, bold = T, color = "white", background = "#0073e6") %>%
  row_spec(1:nrow(combined_chems), background = "#f2f2f2") %>%
  column_spec(1, bold = T) %>%
  add_header_above(c("Chemical Information" = 3))
```

Let's look for ones where use is different between the two states.

```{r}
#| warning: false
#| message: false
combined_chems <- full_join(california_chems, florida_chems, by = "name") %>%
  filter(california_count != florida_count) %>%  # Keep only rows where counts differ
  arrange(desc(abs(california_count - florida_count)))  # Sort by difference in counts

# Step 5: Create the table with kableExtra styling
kable(combined_chems, 
      col.names = c("Chemical", "California Count", "Florida Count"), 
      caption = "Chemicals with Different Counts in California and Florida", 
      align = "lcc") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = F, 
    position = "center") %>%
  row_spec(0, bold = T, color = "white", background = "#0073e6") %>%
  row_spec(1:nrow(combined_chems), background = "#f2f2f2") %>%
  column_spec(1, bold = T) %>%
  add_header_above(c("Chemical Information" = 3))


```

Let's visualize this difference

```{r}
#| warning: false
#| message: false
ggplot(combined_chems, aes(x = reorder(name, california_count - florida_count), y = california_count - florida_count, fill = california_count - florida_count > 0)) +
  geom_bar(stat = "identity", color = "black") +
  coord_flip() +
  labs(title = "Difference in Chemical Usage (California - Florida)",
       x = "Chemical",
       y = "Difference in Count",
       fill = "Higher in") +
  theme_minimal() +
  theme(legend.position = "top", text = element_text(size = 12)) +
  scale_fill_manual(values = c("red", "blue"), labels = c("Florida", "California")) +
  geom_text(aes(label = california_count - florida_count), hjust = -0.2, size = 4)

```


```{r}
#some i picked from WHO pesticides pdf table 3
moderately_dangerous <- c('PYRETHRINS ', 'FLUTRIAFOL ', "FLUPYRADIFURONE ", "FENPYROXIMATE ", "FENPROPATHRIN ", "DIAZINON ")
slightly_dangerous <- c("SULFUR ", "SPINOSAD ", "PENTHIOPYRAD ", "FLUXAPYROXAD ", "CYFLUMETOFEN ", "BORAX DECAHYDRATE ", "BACILLUS SUBTILIS ", "GLYPHOSATE ISO. SALT ")

combined_chems <- combined_chems %>%
  mutate(Category = case_when(
    name %in% moderately_dangerous ~ "Moderately Dangerous",
    name %in% slightly_dangerous ~ "Slightly Dangerous",
    TRUE ~ "Other"
  )) %>%
  filter(Category != "Other")  # Keep only chemicals in defined categories

# Summarize total counts per category per state
dangerous_chems_summary <- combined_chems %>%
  group_by(Category) %>%
  summarise(
    California_Total = sum(california_count, na.rm = TRUE),
    Florida_Total = sum(florida_count, na.rm = TRUE)
  ) %>%
  arrange(desc(California_Total - Florida_Total))

# Create the formatted table
kable(dangerous_chems_summary, 
      col.names = c("Chemical Type", "California Count", "Florida Count"),
      caption = "Comparison of Dangerous Chemicals in California and Florida",
      align = "lcc") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
                full_width = F, position = "center") %>%
  row_spec(0, bold = T, color = "white", background = "#0073e6") %>%
  column_spec(1, bold = T)

dangerous_chems_long <- dangerous_chems_summary %>%
  pivot_longer(cols = c(California_Total, Florida_Total), 
               names_to = "State", 
               values_to = "Count") %>%
  mutate(State = recode(State, "California_Total" = "California", "Florida_Total" = "Florida")) # Ensure proper state names

# Visualization
ggplot(dangerous_chems_long, aes(x = Category, y = Count, fill = State)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Dangerous Chemical Usage",
       x = "Chemical Type",
       y = "Count",
       fill = "State") +
  theme_minimal() +
  scale_fill_manual(values = c("California" = "blue", "Florida" = "red")) # Ensure labels match `State` column

```


## Let's look at organic vs non-organic strawberries

```{r}
#| warning: false
#| message: false
chem_categories_c$Value <- as.numeric(chem_categories_c$Value)

organic_data <- chem_categories_c %>%
  filter(is.na(name))

pesticide_data <- chem_categories_c %>%
  filter(!is.na(name))

organic_data$ChemicalType <- "Organic"
pesticide_data$ChemicalType <- "Pesticide"

combined_data <- bind_rows(organic_data, pesticide_data)

# Plot the histogram with the new 'ChemicalType' variable
ggplot(combined_data, aes(x = ChemicalType, y = Value, fill = ChemicalType)) +
  geom_boxplot(alpha = 0.7, outlier.color = "black", outlier.shape = 16) +
  labs(
    title = "Comparison of Organic vs Pesticide Chemicals",
    x = "Chemical Type",
    y = "Value"
  ) +
  theme_minimal() +
  scale_fill_manual(name = "Chemical Type", values = c("Organic" = "darkgreen", "Pesticide" = "skyblue"))

```


Let's look at Florida vs California. We already saw that California uses more chemicals than Florida, so it would be interesting to see which state makes more money on strawberries. 
```{r}
#| warning: false
#| message: false
straw_sur_c$Item <- trimws(straw_sur_c$Item)
straw_sur_f$Item <- trimws(straw_sur_f$Item)

unique_items <- intersect(unique(straw_sur_c$Item), unique(straw_sur_f$Item))

for (unit in unique_items) {
  
  filtered_c <- straw_sur_c %>%
    filter(Item == unit) %>%
    mutate(Value = as.numeric(Value)) %>%
    drop_na(Value)
  
  filtered_f <- straw_sur_f %>%
    filter(Item == unit) %>%
    mutate(Value = as.numeric(Value)) %>%
    drop_na(Value)
  
  combined_data <- bind_rows(
    filtered_c %>% mutate(State = "California"),
    filtered_f %>% mutate(State = "Florida")
  )
  
  if (nrow(combined_data) > 0) {
    print(
      ggplot(combined_data, aes(x = State, y = Value, fill = State)) +
        geom_boxplot(alpha = 0.7, outlier.color = "black", outlier.shape = 16) +
        labs(
          title = paste("Comparison of", unit, "Between California & Florida"),
          x = "State",
          y = unit
        ) +
        theme_minimal() +
        scale_fill_manual(values = c("California" = "blue", "Florida" = "red"))
    )
  }
}

```
