---
title: "Strawberries"
author: "Elizabeth Rosen"
format: html
editor: visual
---

# Strawberries

## Let's explore chemical usage between Florida and California strawberry fields and see how that contributes to their overall success in the strawberry business using USDA data.

### Loading in the data and cleaning it up to get the census and survey data. Most of the code for loading in the data is from lecture and is visible in the source code for this document.

```{r}
#| label: load libraries
#| warning: false
#| message: false
#| echo: false
library(knitr)  
library(kableExtra)
library(tidyverse)
library(stringr)
```

```{r}
#| label: read data - glimpse 
#| warning: false
#| message: false
#| echo: false

strawberry <- read_csv("strawb_mar6.csv", 
                       col_names = TRUE,
                       show_col_types = FALSE)

source("my_functions.R")

```

Examine the data. How is it organized?

```{r}
#| warning: false
#| message: false
#| echo: false
strawb <- strawberry |> drop_one_value_col()

```

```{r}
#| label: explore strawb data
#| warning: false
#| message: false
#| echo: false
# assume data is a tibble
# n_show is the number of rows to show



show_unique <- function(data, nrows=10 ){
  # make a tibble items to hold the data to show
  # browser()
  a <- nrows * dim(data)[2]  # number of cells in items
  items <- rep(" ", a) # items will coerce everything to char
  dim(items) <- c(nrows ,dim(data)[2]) # shape items
  items <- as_tibble(items)
  colnames(items) <- colnames(data)
  # browser()
  for(i in 1:dim(data)[2]){

    col_items <- unique(data[,i])
    # row_ex is the number of rows needed 
    # to make the column length conformable with items
    row_ex <- nrows - dim(col_items)[1] 
    if(row_ex >= 0){
      ex_rows <- tibble(rep(" ",row_ex))
      colnames(ex_rows) <- colnames(col_items)
      col_add <- rbind2(col_items, ex_rows)
      
    } else if(row_ex < 0){
      col_add <- col_items[1:10,]
      
    }

    items[,i] <- col_add

  }
  
  return(items)
}

## test <- show_unique(strawb, 10)


```

```{r}
#| label: split strawb into census and survey pieces
#| warning: FALSE
#| message: FALSE
#| echo: FALSE
#| results: FALSE
strw_census <- strawb |> filter(Program == "CENSUS")

strw_survey <- strawb |> filter(Program == "SURVEY")

nrow(strawb) == (nrow(strw_census) + nrow(strw_survey))

```

```{r}
#| label: examine Census and survey tibbles
#| warning: false
#| message: false
#| echo: false
#| results: FALSE
s_census <- strw_census |> drop_one_value_col(prt_val = TRUE)

s_survey <- strw_survey |> drop_one_value_col(prt_val = TRUE)


unique_sur <- s_survey |> show_unique(nrows = 10)

unique_cen <- s_census |> show_unique(nrows = 10)


strw_census <- s_census |> select(-`State ANSI`)

strw_survey <- s_survey |> select(-`State ANSI`, -`Week Ending`, -Period)

#rm(s_census, s_survey, strawberry, strawb, items)


```

```{r}
#| label: work on Census columns
#| warning: false
#| message: false
#| echo: false
commod <- strw_census$Commodity |> unique()

#### split Data Item

strw_census <- strw_census |>
  separate_wider_delim(  cols = Commodity,
                         delim = ",",
                         names = c("INCOME", 
                                   "NET CASH FARM",
                                   "STRAW"
                                               ),
                         too_many = "error",
                         names_sep = " ",
                         too_few = "align_start"
  )



inc <- strw_census$Fruit |> unique()

strw_census <- strw_census |>
  separate_wider_delim(  cols = Fruit,
                         delim = ",",
                         names = c("INCOME", 
                                   "STRAWB"
                                               ),
                         too_many = "error",
                         too_few = "align_start"
  )

```

```{r}
#| label: Assignment shortner results
#| warning: false
#| message: false
#| echo: false
straw_cen_f <- strw_census |> filter(State == "FLORIDA")

straw_sur_f <- strw_survey |> filter(State == "FLORIDA")
straw_cen_c <- strw_census |> filter(State == "CALIFORNIA")
straw_sur_c <- strw_survey |> filter(State == "CALIFORNIA")

rm(strw_census, strw_survey, unique_cen, unique_sur)

```

## Chemical usage

### Let us explore how chemicals are used differently between the two states!

This code is all my own. I used this <https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html> for help with the kable extra tables. I also looked up the specific color names so I could experiment with making the table colors look pretty.

```{r}
#| warning: false
#| message: false
#| echo: FALSE

chem_categories_c <- straw_sur_c |> separate_wider_delim(cols='Domain Category',
                                                            delim=': ',
                                                            names=c('domain','chemical name'),
                                                            too_few='align_start',
                                                            too_many = 'error') %>% select(-domain)
chem_categories_c <- chem_categories_c |> separate_wider_delim(cols='chemical name',
                                                               delim='=',
                                                               names = c('name', 'number'),
                                                               too_few='align_start',
                                                               too_many='error') |> separate_wider_delim(cols='name', delim='(', names=c('extra', 'name'), too_few='align_start', too_many='error') %>% select(-extra)
unique_values_c <- unique(chem_categories_c$`name`)

chem_categories_f <- straw_sur_f |> separate_wider_delim(cols='Domain Category',
                                                            delim=': ',
                                                            names=c('domain','chemical name'),
                                                            too_few='align_start',
                                                            too_many = 'error') %>% select(-domain)
chem_categories_f <- chem_categories_f |> separate_wider_delim(cols='chemical name',
                                                               delim='=',
                                                               names = c('name', 'number'),
                                                               too_few='align_start',
                                                               too_many='error') |> separate_wider_delim(cols='name', delim='(', names=c('extra', 'name'), too_few='align_start', too_many='merge') %>% select(-extra)
unique_values_f <- unique(chem_categories_f$`name`)

```

### Let's look at the chemicals used in California and in Florida.

```{r}
#| warning: false
#| message: false
#| echo: FALSE

chems_c <- chem_categories_c %>%
  filter(name != "TOTAL)") %>%
  group_by(name) %>%
  summarise(n = n()) 

chems_f <- chem_categories_f %>%
  filter(name != "TOTAL)") %>%
  group_by(name) %>%
  summarise(n = n()) 

kable(chems_c, 
      col.names = c("Chemical", "Count"), 
      caption = "Count of Chemicals Used in California", 
      align = "lcc") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = F, 
    position = "center") %>%
  row_spec(0, bold = T, color = "white", background = "powderblue") %>%
  row_spec(1:nrow(chems_c), background = "grey97") %>%
  column_spec(1, bold = T) %>%
  add_header_above(c("Chemical Information" = 2))

kable(chems_f, 
      col.names = c("Chemical", "Count"), 
      caption = "Count of Chemicals Used in Florida", 
      align = "lcc") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = F, 
    position = "center") %>%
  row_spec(0, bold = T, color = "white", background = "pink") %>%
  row_spec(1:nrow(chems_f), background = "grey97") %>%
  column_spec(1, bold = T) %>%
  add_header_above(c("Chemical Information" = 2))
```

## Which ones do they have in common? Let's focus on these.

```{r}
#| warning: false
#| message: false
#| echo: FALSE
common_chems <- intersect(chems_c$name, chems_f$name)

california_chems <- chems_c %>%
  filter(name %in% common_chems) %>%
  group_by(name) %>%
  summarise(california_count = sum(n), .groups = "drop")  

florida_chems <- chems_f %>%
  filter(name %in% common_chems) %>%
  group_by(name) %>%
  summarise(florida_count = sum(n), .groups = "drop")  

combined_chems <- full_join(california_chems, florida_chems, by = "name") %>%
  arrange(desc(california_count + florida_count))  


kable(combined_chems, 
      col.names = c("Chemical", "California Count", "Florida Count"), 
      caption = "Counts of Common Chemicals in California and Florida", 
      align = "lcc") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = F, 
    position = "center") %>%
  row_spec(0, bold = T, color = "white", background = "powderblue") %>%
  row_spec(1:nrow(combined_chems), background = "grey97") %>%
  column_spec(1, bold = T) %>%
  add_header_above(c("Chemical Information" = 3))
```

## Let's look for ones where use is different between the two states.

```{r}
#| warning: false
#| message: false
#| echo: FALSE

combined_chems <- full_join(california_chems, florida_chems, by = "name") %>%
  filter(california_count != florida_count) %>%  
  arrange(desc(abs(california_count - florida_count)))  

kable(combined_chems, 
      col.names = c("Chemical", "California Count", "Florida Count"), 
      caption = "Chemicals with Different Counts in California and Florida", 
      align = "lcc") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = F, 
    position = "center") %>%
  row_spec(0, bold = T, color = "white", background = "pink") %>%
  row_spec(1:nrow(combined_chems), background = "grey97") %>%
  column_spec(1, bold = T) %>%
  add_header_above(c("Chemical Information" = 3))


```

## Let's visualize this difference

```{r}
#| warning: false
#| message: false
#| echo: FALSE
ggplot(combined_chems, aes(x = reorder(name, california_count - florida_count), y = california_count - florida_count, fill = california_count - florida_count > 0)) +
  geom_bar(stat = "identity", color = "black") +
  coord_flip() +
  labs(title = "Difference in Chemical Usage (California - Florida)",
       x = "Chemical",
       y = "Difference in Count",
       fill = "Higher in") +
  theme_minimal() +
  theme(legend.position = "top", text = element_text(size = 12)) +
  scale_fill_manual(values = c("pink", "powderblue"), labels = c("Florida", "California")) +
  geom_text(aes(label = california_count - florida_count), hjust = -0.2, size = 4)

```

### I selected the chemicals from this table that are classified as either slightly or moderately dangerous according to the WHO to see which state uses more dangerous chemicals.

```{r}
#| warning: false
#| message: false
#| echo: FALSE
#some i picked from WHO pesticides pdf table 3
moderately_dangerous <- c('PYRETHRINS ', 'FLUTRIAFOL ', "FLUPYRADIFURONE ", "FENPYROXIMATE ", "FENPROPATHRIN ", "DIAZINON ")
slightly_dangerous <- c("SULFUR ", "SPINOSAD ", "PENTHIOPYRAD ", "FLUXAPYROXAD ", "CYFLUMETOFEN ", "BORAX DECAHYDRATE ", "BACILLUS SUBTILIS ", "GLYPHOSATE ISO. SALT ")

combined_chems <- combined_chems %>%
  mutate(Category = case_when(
    name %in% moderately_dangerous ~ "Moderately Dangerous",
    name %in% slightly_dangerous ~ "Slightly Dangerous",
    TRUE ~ "Other"
  )) %>%
  filter(Category != "Other")  

dangerous_chems_summary <- combined_chems %>%
  group_by(Category) %>%
  summarise(
    California_Total = sum(california_count, na.rm = TRUE),
    Florida_Total = sum(florida_count, na.rm = TRUE)
  ) %>%
  arrange(desc(California_Total - Florida_Total))

kable(dangerous_chems_summary, 
      col.names = c("Chemical Type", "California Count", "Florida Count"),
      caption = "Comparison of Dangerous Chemicals in California and Florida",
      align = "lcc") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
                full_width = F, position = "center") %>%
  row_spec(0, bold = T, color = "white", background = "lavender") %>%
  column_spec(1, bold = T)

dangerous_chems_long <- dangerous_chems_summary %>%
  pivot_longer(cols = c(California_Total, Florida_Total), 
               names_to = "State", 
               values_to = "Count") %>%
  mutate(State = recode(State, "California_Total" = "California", "Florida_Total" = "Florida")) 

ggplot(dangerous_chems_long, aes(x = Category, y = Count, fill = State)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Dangerous Chemical Usage",
       x = "Chemical Type",
       y = "Count",
       fill = "State") +
  theme_minimal() +
  scale_fill_manual(values = c("California" = "powderblue", "Florida" = "pink")) 

```

# Let's look at organic vs non-organic strawberries

```{r}
#| warning: false
#| message: false
#| echo: FALSE
chem_categories_c$Value <- as.numeric(chem_categories_c$Value)

organic_data <- chem_categories_c %>%
  filter(is.na(name))

pesticide_data <- chem_categories_c %>%
  filter(!is.na(name))

organic_data$ChemicalType <- "Organic"
pesticide_data$ChemicalType <- "Pesticide"

combined_data <- bind_rows(organic_data, pesticide_data)

ggplot(combined_data, aes(x = ChemicalType, y = Value, fill = ChemicalType)) +
  geom_boxplot(alpha = 0.7, outlier.color = "black", outlier.shape = 16) +
  labs(
    title = "Comparison of Organic vs Pesticide Chemicals",
    x = "Chemical Type",
    y = "Value"
  ) +
  theme_minimal() +
  scale_fill_manual(name = "Chemical Type", values = c("Organic" = "olivedrab3", "Pesticide" = "coral"))

```

## Let's look at Florida vs California. We already saw that California uses more chemicals than Florida, so it would be interesting to see which state makes more money on strawberries.

```{r}
#| warning: false
#| message: false
#| echo: false
straw_sur_c$Item <- trimws(straw_sur_c$Item)
straw_sur_f$Item <- trimws(straw_sur_f$Item)

unique_items <- intersect(unique(straw_sur_c$Item), unique(straw_sur_f$Item))

for (unit in unique_items) {
  
  filtered_c <- straw_sur_c %>%
    filter(Item == unit) %>%
    mutate(Value = as.numeric(Value)) %>%
    drop_na(Value)
  
  filtered_f <- straw_sur_f %>%
    filter(Item == unit) %>%
    mutate(Value = as.numeric(Value)) %>%
    drop_na(Value)
  
  combined_data <- bind_rows(
    filtered_c %>% mutate(State = "California"),
    filtered_f %>% mutate(State = "Florida")
  )
  
  if (nrow(combined_data) > 0) {
    print(
      ggplot(combined_data, aes(x = State, y = Value, fill = State)) +
        geom_boxplot(alpha = 0.7, outlier.color = "black", outlier.shape = 16) +
        labs(
          title = paste("Comparison of", unit, "Between California & Florida"),
          x = "State",
          y = unit
        ) +
        theme_minimal() +
        scale_fill_manual(values = c("California" = "powderblue", "Florida" = "pink"))
    )
  }
}

```

Based upon these graphs, it seems that California sells more pounds of strawberries than Florida, but Florida sells more in pounds per acre of fields. I would conclude that Florida has a more lucrative strawberry business for the size of their strawberry fields, and that their reduced use of dangerous chemicals could be beneficial to their business.
